Gastro-intestinal transmission of _Clostridioides difficile_ and antimicrobial-resistant bacteria in healthcare settings -- calibration of a simulation model

### 2025-12-03 paper/paper.qmd
ABM math updates

---
title: Gastro-intestinal transmission of _Clostridioides difficile_ and antimicrobial-resistant bacteria in healthcare settings -- calibration of an agent-based model
authors:
  - name: Alessandro Gerada
    affiliation: University of Liverpool
    roles: writing
    corresponding: true
bibliography: references.bib
csl: https://raw.githubusercontent.com/citation-style-language/styles/refs/heads/master/elsevier-vancouver.csl
filters:
  - acronyms
acronyms:
  keys:
  - shortname: CDI
    longname: _Clostridium difficile_ infection
    markdown: longname
  - shortname: VRE
    longname: vancomycin-resistant _Enterococcus_
    markdown: longname
  - shortname: ESBL
    longname: extended-spectrum beta-lactamase
  - shortname: CPE
    longname: carbapenemase-producing Enterobacterales
  - shortname: WHO
    longname: World Health Organization
  - shortname: IPC
    longname: infection prevention and control
  - shortname: ABM
    longname: agent-based model
  - shortname: MRSA
    longname: methicillin-resistant _Staphylococcus aureus_
    markdown: longname
  - shortname: SARS-CoV-2
    longname: severe acute respiratory syndrome coronavirus 2
  - shortname: ABC
    longname: approximate Bayesian computation
  - shortname: IQR
    longname: interquartile range
  - shortname: FIFO
    longname: first-in, first-out
  - shortname: SD
    longname: system dynamics
  insert_loa: false
  insert_links: false
---

@@

Since the simulation's target diseases were associated with long term colonisation and prolonged risk of transmission to other patients, it was assumed that infection was not reversible during the simulation, i.e., that infected patients remained infected until discharge. At each time increment, the probability of a non-infected patient $i$ acquiring infection (i.e., transitioning from a non-infected to infected state)---$P_i(t)$---depended on: 1) a baseline risk of *de novo* infection; and 2) the usage of contaminated toilets:

@@

@@

-   $P_i(t)$ is the risk of patient $i$ acquiring infection at time $t$,

@@

-   $m$ is the increased (or decreased) relative risk of infection due to antimicrobial therapy.

@@

$$
X_\text{de novo, i}(t) \sim \text{Bernoulli}(P_{\text{de novo}, i}(t)),
$$

where $X_{\text{de novo}, i}$ = 1 if patient $i$ becomes infected *de novo* at time $t$, and 0 otherwise.

Further Bernoulli trials were conducted to determine if a patient became infected from a contaminated toilet at each time step $t$:

$$
X_{\text{toilet}, i}(t) \sim \text{Bernoulli}(P_{\text{toilet}, i}(t)),
$$

where $X_{\text{toilet}, i}$ = 1 if patient $i$ becomes infected from a contaminated toilet at time $t$, and 0 otherwise.

The results of the trials were combined to determine the outcome for patient $p$ at time $t$; the patient became infected if either trial was successful:

$$
X_i(t) = \text{Max}(X_{\text{de novo}, i}, X_{\text{toilet}, i}),
$$

where $X_i(t)$ = 1 if patient $i$ becomes infected at time $t$, and 0 otherwise.

@@

Infected patients were moved to single-occupancy rooms (side rooms) after infection. To capture delays in moving patients into side rooms (which can be due to delays in diagnosis, and/or lack of available side rooms), the simulation included an optional parameter that samples the delay from a Poisson distribution with a configurable mean (in days). Once the infected patient has completed their delay, they became eligible to move to the next available side room. To implement this, the simulation used a \acr{FIFO} queue to track infected patients waiting for a side room. While direct patient swaps were possible (i.e., swapping a non-infected patient in a side room with an infected patient waiting for a side room), more complex swaps involving multiple patients were not implemented.

Each patient's admission duration was sampled from a Poisson distribution with a configurable mean (in days). Upon discharge, patients were removed from the simulation, and replaced with new admissions to maintain 100% occupancy. New admissions had a probability of being infected based on the community infection rate parameter.

>>>>

---
title: Gastro-intestinal transmission of _Clostridioides difficile_ and antimicrobial-resistant bacteria in healthcare settings -- calibration of a simulation model
authors:
  - name: Alessandro Gerada
    affiliation: University of Liverpool
    roles: writing
    corresponding: true
bibliography: references.bib
csl: https://raw.githubusercontent.com/citation-style-language/styles/refs/heads/master/elsevier-vancouver.csl
filters:
  - acronyms
acronyms:
  keys:
  - shortname: CDI
    longname: _Clostridium difficile_ infection
    markdown: longname
  - shortname: VRE
    longname: vancomycin-resistant _Enterococcus_
    markdown: longname
  - shortname: ESBL
    longname: extended-spectrum beta-lactamase
  - shortname: CPE
    longname: carbapenemase-producing Enterobacterales
  - shortname: WHO
    longname: World Health Organization
  - shortname: IPC
    longname: infection prevention and control
  - shortname: ABM
    longname: agent-based model
  - shortname: MRSA
    longname: methicillin-resistant _Staphylococcus aureus_
    markdown: longname
  - shortname: SARS-CoV-2
    longname: severe acute respiratory syndrome coronavirus 2
  - shortname: ABC
    longname: approximate Bayesian computation
  - shortname: IQR
    longname: interquartile range
  - shortname: FIFO
    longname: first-in, first-out
  - shortname: SD
    longname: system dynamics
  insert_loa: false
  insert_links: false
---

@@

On admission, each patient had a probability $\lambda$ of being infected, representing the community infection rate. The simulation determined the infection status of each patient on admission using Bernoulli trials:

$$
X_{\text{community}, i} \sim \text{Bernoulli}(\lambda),
$$

where $X_{\text{community}, i} = 1$ if patient $i$ is infected on admission, and 0 otherwise.

Since the simulation's target diseases were associated with long term colonisation and prolonged risk of transmission to other patients, it was assumed that infection was not reversible during the simulation, i.e., that infected patients remained infected until discharge. At each time increment, the probability of a non-infected patient $i$ acquiring infection (i.e., transitioning from a non-infected to infected state)---$P_i(t)$---depended on: 1) a baseline risk of *de novo* infection; and 2) the usage of contaminated toilets. The probability of infection through these two routes---$P_{\text{de novo}, i}(t)$ and $P_{\text{toilet}, i}(t)$, respectively---were independent; therefore, the probability of patient $i$ acquiring infection at time $t$ was given by:

@@

* $(1- P_{\text{de novo}, i}(t))$ is the probability of patient $i$ not acquiring infection *de novo* at time $t$,

* $(1 - P_{\text{toilet}, i}(t))$ is the probability of patient $i$ not acquiring infection from a contaminated toilet at time $t$, and

* $[(1 - P_{\text{de novo}, i}(t)) \cdot (1 - P_{\text{toilet}, i}(t))]$ is the probability of patient $i$ not acquiring infection at time $t$.

The individual probabilities were defined as:

@@

@@

-   $m$ is the increased (or decreased) relative risk of infection due to antimicrobial therapy, where $m>0$.

@@

$$
X_{\text{hospital}, i}(t) \sim \text{Bernoulli}(P_i(t))
$$

where $X_{\text{hospital}, i}(t) = 1$ if patient $i$ acquires infection during their hospital stay at time $t$, and 0 otherwise.

@@

Infected patients were moved to single-occupancy rooms (side rooms) after infection. To capture delays in moving patients into side rooms (which can be due to delays in diagnosis, and/or lack of available side rooms), the simulation included an optional parameter that samples the delay from a Poisson distribution with a configurable mean (in days). Once the infected patient completed their delay, they became eligible to move to the next available side room. To implement this, the simulation used a \acr{FIFO} queue to track infected patients waiting for a side room. While direct patient swaps were possible (i.e., swapping a non-infected patient in a side room with an infected patient waiting for a side room), more complex swaps involving multiple patients were not implemented.

Each patient's admission duration was sampled from a Poisson distribution with a configurable mean (in days). Upon discharge, patients were removed from the simulation; each patient was then replaced with new patient admission. Although real hospitals do not operate at 100% occupancy at all times, this assumption was made to simplify the model.
